What is an Ancestral Recombination Graph?

### 2022-02-24 paper.tex
Rename the simplification figure
Change fig 2 to compare node-vs-edge annotations

\begin{figure}
\centering
\begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

@@

\node [nodelabel,anchor=north west] at ($(-6.5,5)$) {
\begin{tabular}{c|c}
\multicolumn{2}{c}{Breakpoint}\\
\hline
0 & $\infty$ \\
1 & $\infty$ \\
2 & $\infty$ \\
3 & $x$ \\
4 & $\infty$ \\
5 & $\infty$ \\
6 & $\infty$ \\
\end{tabular}};

@@

\end{tikzpicture}
&
\begin{tikzpicture}[x=5mm,y=5mm,node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] at (0, 0) {};
\node (s1) [greynode] at (3, 0) {};
\node (s2) [greynode] at (6, 0) {};
\node (s3) [greynode] at (2, 1) {};
\node (s4) [greynode] at (4, 1) {};
\node (s5) [greynode] at (1, 2) {};
\node (s6) [greynode] at (5, 3) {};
\node (s7) [greynode] at (3, 4) {};

\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s3/3} \node[nodelabel,anchor=east] at (\u) {\lab};
\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=west] at (\u) {\lab};
\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s6/6} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s7/7} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -| (s3);
\draw (s1) -| (s4) |- (s6);
\draw (s3) |- (s5);
\draw (s0) |- (s5);
\draw (s2) |- (s6);
\draw (s5) |- (s7);
\draw (s6) |- (s7);

\node [nodelabel,anchor=north west] at ($(-5,5)$) {
\begin{tabular}{ll}
Edges\\
\hline
$(0, 5)$ & $(0, L]$ \\
$(1, 3)$ & $(0, x]$ \\
$(1, 4)$ & $(x, L]$ \\
$(2, 6)$ & $(0, L]$ \\
$(3, 5)$ & $(0, L]$ \\
$(4, 6)$ & $(0, L]$ \\
$(5, 7)$ & $(0, L]$ \\
$(6, 7)$ & $(0, L]$ \\
\end{tabular}};

\end{tikzpicture}
\\
(A) & (B) \\
\end{tabular}
\caption{\label{fig-arg-data-structure}
Encoding an ARG. (A) The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoints with nodes. This encoding
is limited in the ancestral patterns that can be described and ambiguous.
[TODO clarify that the function is the same things as having two
types of nodes]
(B) Equating graph nodes with \emph{genomes} and attaching the
interval of genome inherited by the child node with edges removes
this ambiguity and can describe any pattern of genetic inheritance.
The most important change between the two encodings is to represent
a recombination event (node 3 in A) by two parent genomes
(nodes 3 and 4 in B) and explicitly storing the interval
that the child inherits from each parent along the two corresponding
edges. Thus, child $1$ in inherits $(0, x]$ from $3$
and $(x, L)$ from $4$
(for recombination breakpoint $x$ and a sequence of length $L$).
\textcolor{red}{This is very much a draft; let's see how it works with
 the text and other figures and tidy it up in a later revision}
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ARG_edge_annotations}
\caption{\label{fig-ancestry-resolution}
Implicit vs explicit edge annotations. (a) An gARG equivalent to the encoding
described by Griffiths can be created by setting all edges to span the full
genome, apart from edges between two recombining nodes and their recombinant
child (edge spans in bold). (b) The sample-resolved gARG, where the Hudson
simplification algorithm has shortened edge spans so that they encapsulate only
the upwards transmission of genetic material from the samples A, B, and C;
plotted line widths are proportional to the genomic span of each edge. The
example ARG used here is taken from \citet[][fig. 1]{wiuf1999recombination}.
}
\end{figure}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-simplification}
ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
model. Note the complex full-sib relationship between B and C; this
could not be directly represented in a Griffiths-like EARG encoding
(see section XXX).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds).
(C) Remove nodes that are unary in all local trees.
(D) Rewrite edges to skip any unary nodes in local trees.
(E) The local trees for ARG (A).
(F) The local trees for ARG (D).
}
\end{figure}

>>>>

\begin{figure}
\centering
% \begin{tabular}{cc}
\begin{tikzpicture}[x=5mm, y=5mm, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

@@

\node [nodelabel,anchor=north west] at ($(-8.5,5)$) {
\begin{tabular}{c|c}
% \multicolumn{2}{c}{Breakpoint}\\
node & breakpoint\\
\hline
0 & $\varnothing$ \\
1 & $\varnothing$ \\
2 & $\varnothing$ \\
3 & $x$ \\
4 & $\varnothing$ \\
5 & $\varnothing$ \\
6 & $\varnothing$ \\
\end{tabular}};

@@

\node [nodelabel,anchor=north west] at ($(6,5)$) {
\begin{tabular}{ll}
Edges\\
\hline
$(0, 4)$ & $(0, L]$ \\
$(1, 3)$ & $(0, L]$ \\
$(2, 5)$ & $(0, L]$ \\
$(3, 4)$ & $(0, x]$ \\
$(3, 5)$ & $(x, L]$ \\
$(4, 6)$ & $(0, L]$ \\
$(5, 6)$ & $(0, L]$ \\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-arg-data-structure}
Node vs edge breakpoint annotations.
The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoint information with \emph{nodes} (left).
This encoding
is limited in the pattern of inheritance that can be described
and requires that edges be ordered to avoid ambiguity.
[TODO clarify that the function is the same things as having two
types of nodes]
The alternative is to associate breakpoint annotations
with \emph{edges} (right).
Associating the interval of genome inherited by the child node
from the parent with the corresponding edge, allows
any pattern genetic inheritance to be described and does
not required edges to be ordered.
}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/ancestry-resolution}
\caption{\label{fig-ancestry-resolution}
Implicit vs explicit edge annotations. (a) An gARG equivalent to the encoding
described by Griffiths can be created by setting all edges to span the full
genome, apart from edges between two recombining nodes and their recombinant
child (edge spans in bold). (b) The sample-resolved gARG, where the Hudson
simplification algorithm has shortened edge spans so that they encapsulate only
the upwards transmission of genetic material from the samples A, B, and C;
plotted line widths are proportional to the genomic span of each edge. The
example ARG used here is taken from \citet[][fig. 1]{wiuf1999recombination}.
}
\end{figure}

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/simplification}
\caption{\label{fig-simplification}
ARG simplification. (A) An ARG simulated from a diploid Wright-Fisher
model. Note the complex full-sib relationship between B and C; this
could not be directly represented in a Griffiths-like EARG encoding
(see section XXX).
(B) Simplified to remove all
% trying out this terminology - can define in the text
1-connected graph components (e.g., diamonds).
(C) Remove nodes that are unary in all local trees.
(D) Rewrite edges to skip any unary nodes in local trees.
(E) The local trees for ARG (A).
(F) The local trees for ARG (D).
}
\end{figure}
